The Multi-Stencil Language

### 2016-05-30 ARTICLES/IJPP/article/formalism.tex
Formalism

Some approaches to numerically solve partial differential equations (PDEs) are based on direct iterative methods (\textit{e.g.} finite difference, finite volume or finite element methods).
They approximate the solution through a discretized process where the continuous time and space domains are discretized and numerical computations are iteratively (time discretization) applied onto a mesh (space discretization).
%In other words, the PDEs are transformed to a set of numerical computations applied at each time step on elements of the discretized space domain.
While the computations can have various forms, we focus on three categories.
\emph{Stencil} computations involve access to neighbor values only (the concept of neighborhood depending on the space discretization used).
\emph{Local} computations depend on the computed location only (this can be seen as a stencil of size one).
Finally, \emph{reductions} enable to transform values mapped on the mesh to a single scalar value.

@@

\begin{mydef}[Definitions (time)]
The finite sequence $T: (t_n)_{n\in\llbracket 0, Tmax \rrbracket}$ represents the discretization of the continuous time domain $\mathcal{T}=\mathbb{R}$.
To each discrete time step $n\in\llbracket 0, Tmax \rrbracket$, it associates a time value $t_n\in\mathcal{T}$.
\end{mydef}

@@

\begin{mydefs}[Definitions (quantity)]
\item $\mathbb{V}=\Delta\cup\mathcal{S}$ is the set of \emph{variables} or \emph{quantities}.
\item $\Delta$ are the \textit{mesh variables} that to each entity of a group and time-step associates a value $\delta: \mathcal{G}\times T\mapsto \mathcal{V}_\delta$ where $V_{\delta}$ is a value type.
\item The group of entities a variable is mapped on is denoted $entity(\delta)=\mathcal{G}$.
\item $\mathcal{S}$ are the \textit{scalar variables} that to each time-step associates a value $s: T\mapsto \mathcal{V}_\delta$ where $V_{\delta}$ is a value type.
\item Amongst the scalar variables is one specific variable $conv\in\mathcal{S}$, the convergence criteria whose value is $0$ except at the last iteration where it is 1 $\forall t\in \llbracket 0, Tmax-1 \rrbracket conv(t)=1, conv(Tmax)=1$.
\end{mydefs}

%For example, the simplest $conv$ is typically to have two scalars: $t$ the time, and a constant number of iterations denoted $it$.
%At each time step $\Delta t=1$, and $conv$ returns the boolean expression $t<it$.

@@

This section has presented the general formalism of what is a mesh, its entities and groups of entities, of how the time is discretized, and finally of what is a scalar and a quantity.
The language presented in this paper, the Multi-Stencil Language, offers a simple way to describe a numerical simulation while being able to automatically extract parallelism.
Its aim is to be mesh-agnostic, thus to not have details about the topology of the mesh which is an implementation problem.
As a result the details about the mesh topology are not needed in the rest of this paper.

@@

In this section are considered four different types of kernel computations, stencil kernels, boundary kernels, local kernels, and reduction kernels.

\medskip
\begin{mydefs}
\item A computation domain $D$ is a subpart of a mesh entities group, $D \subseteq \mathcal{G} \in \Phi$.
\item The set of computation domains of a numerical simulation is denoted $\mathcal{D}$.
\item A neighborhood $n$ is a function which for a given entity $\phi \in \mathcal{G}_i$, returns a set of $m$ entities in $\mathcal{G}_j$, $n : \mathcal{G}_i \rightarrow \mathcal{G}_j^m$. One can notice that $i = j$ is possible. Most of the time, such a neighborhood is called a \emph{stencil shape}.
\item The set of neihborhood functions in a numerical simulation is denoted $\mathcal{N}$.
\end{mydefs}

\begin{mydef}
A computation kernel $k$ of a numerical simulation is defined as $k(S,R,(w,D),comp)$, where 
\begin{itemize}
\item $S \in \mathcal{S}$ is the set of scalar to read, 
\item $(w,D) \in \Delta \times \mathcal{D}$ is the single data modified by the computation kernel onto a given computation domain. Thus, $w \in \Delta$ and $D \in \mathcal{D}$ is the computation domain on which $w$ is computed, $D \subseteq \mathcal{G}_w=entity(w)$.
\item $R$ is the set of tuples $(r,n)$, where $r \in \Delta$ and $n \in \mathcal{N}$ is a neighborhood function such that $n : \mathcal{G}_w \rightarrow entity(r)^m$. The neighborhood indicates which entities of $r$ will be read in order to compute a single entity of $w$. 
\item $comp$ is the numerical computation which returns a value from a set of $n$ input values, $comp: V^n \rightarrow V$, where $V$ is typically $\mathbb{N}$, $\mathbb{Z}$, or $\mathbb{R}$. $comp$ represents the actual numerical expression which is computed by a kernel.
\end{itemize}
\end{mydef}

At each time iteration, a set of computations is performed. During a computation kernel, it can be considered that a set of old states ($t-1$) of quantities are read ($R$), and that a new state ($t$) of a single quantity is written ($w$). In this paper only explicit numerical schemes are considered, thus if we denote by $identity$ the identity function, we have $w \not\in R$ except if $w=(r,identity)$.

Such a definition of a computation kernel offers a large panel of different computations. For example, the four usual types of computations (stencil, local, boundary and reduction) performed into a simulation can be defined as follow :
\begin{itemize}
\item A kernel computation $k(S,R,(w,D),comp)$ is a \emph{stencil kernel} if $\exists (r,n) \in R$ such that $n \neq identity$ .
\item A \emph{boundary kernel} is a kernel $k(S,R,(w,D),comp)$ where $D$ is a specific computation domain at the border of entities, and which does not intersect with any other computation domain.
\item A kernel computation $k(S,R,(w,D),comp)$ is a \emph{local kernel} if $\forall (r,n) \in R$, $n = identity$.
\item A kernel computation $k(S,R,(w,D),comp)$ is a \emph{reduction kernel} if $w$ is a scalar.%, $D=\emptyset$, $R=\{(r,n)\}$ with $n=entity(r)$, and $comp$ is a reduction operation (usually $+$, $-$, $*$, $max$, $min$ etc.).
\end{itemize}

%In our formalism, it is forbidden for a stencil kernel to find $(r,n) \in R$ for which $r=w$ and $n \neq identity$. Actually, if such a property was permitted, implicit numerical schemes would be needed into the simulation, which involves linear solvers. Such a scheme is not a stencil and is over the scope of this paper. The problem does not happened for local kernels as $\forall (r,n) \in R$, $n = identity$. Computations for which this property is authorized are boundary kernels only. This type of kernel is a particular case as it commputes boundary conditions, which represents what should happened outside the space domain but still impact stencil kernels at the next time step.

@@

\paragraph{\textbf{Summary}} In this subsection have been presented the general formalism of what is a computation kernel and a multi-stencil program. The language presented in this paper, the Multi-Stencil Language, offers a simple way to describe a numerical simulation while being able to automatically extract parallelism. Its aim is to be mesh-agnostic, as previously described, but also to make a clear separation of concerns between the simulation description and its implementation. As a result, and as already described, details about the topology of the mesh, which is an implementation problem, is not used into the language. Moreover, in a computation kernel $k(S,R,(w,D),comp)$, $comp$ is not used by the language.

>>>>

Some approaches to numerically solve partial differential equations (PDEs) are based on direct iterative methods (\textit{e.g.} finite difference, finite volume or finite element methods).
They approximate the solution through a discretized process where the continuous time and space domains are discretized and numerical computations are iteratively (time discretization) applied onto a mesh (space discretization).
%In other words, the PDEs are transformed to a set of numerical computations applied at each time-step on elements of the discretized space domain.
While the computations can have various forms, we focus on three categories.
%TODO[JB] 4 plus tard...
\emph{Stencil} computations involve access to neighbor values only (the concept of neighborhood depending on the space discretization used).
\emph{Local} computations depend on the computed location only (this can be seen as a stencil of size one).
Finally, \emph{reductions} enable to transform values mapped on the mesh to a single scalar value.

@@

\begin{mydef}[Definitions (time)]
The finite sequence $T: (t_n)_{n\in\llbracket 0, Tmax \rrbracket}$ represents the discretization of the continuous time domain $\mathcal{T}=\mathbb{R}$.
To each discrete time-step $n\in\llbracket 0, Tmax \rrbracket$, it associates a time value $t_n\in\mathcal{T}$.
\end{mydef}

@@

\begin{mydefs}[Definitions (quantity)]
\item $\mathbb{V}=\Delta\cup\mathcal{S}$ is the set of \emph{variables} or \emph{quantities}.
\item $\Delta$ are the \textit{mesh variables} that to each entity of a group and time-step associates a value $\delta: \mathcal{G}\times T\mapsto \mathcal{V}_\delta$ where $V_{\delta}$ is a value type.
\item The group of entities a variable is mapped on is denoted $entity(\delta)=\mathcal{G}$.
\item $\mathcal{S}$ are the \textit{scalar variables} that to each time-step associates a value $s: T\mapsto \mathcal{V}_\delta$ where $V_{\delta}$ is a value type.
\item Amongst the scalar variables is one specific variable $conv\in\mathcal{S}$, the convergence criteria whose value is $0$ except at the last step where it is 1 $\forall t\in \llbracket 0, Tmax-1 \rrbracket conv(t)=1, conv(Tmax)=1$.
\end{mydefs}

%For example, the simplest $conv$ is typically to have two scalars: $t$ the time, and a constant number of iterations denoted $it$.
%At each time-step $\Delta t=1$, and $conv$ returns the boolean expression $t<it$.

@@

This section has presented the general formalism of meshes, their entities, groups of entities, time discretization, and finally quantities (mapped on mesh and scalars).
While the details about mesh and their topologies were useful to specify the following concepts, the Multi-Stencil Language presented in this paper is mesh-agnostic.
These details will therefore not be needed in the remaining of the paper.

@@

In this section four different types of kernel computations are considered: stencil kernels, boundary kernels, local kernels, and reduction kernels.

\medskip
\begin{mydefs}
\item A computation domain $D$ is a subpart of a mesh entities group, $D \subseteq \mathcal{G} \in \Phi$.
\item The set of computation domains of a numerical simulation is denoted $\mathcal{D}$.
\item $\mathcal{N}$ are the neighborhood functions $n: \mathcal{G}_i \mapsto \mathcal{G}_j^m$ which for a given entity $\phi \in \mathcal{G}_i$ returns a set of $m$ entities in $\mathcal{G}_j$. One can notice that $i = j$ is possible. Most of the time, such a neighborhood is called a \emph{stencil shape}.
\end{mydefs}

\begin{mydef}
A computation kernel $k$ of a numerical simulation is defined as $k=(S,R,(w,D),comp)$, where
\begin{itemize}
\item $S \in \mathcal{S}$ is the set of scalar to read,
\item $(w,D) \in \Delta \times \mathcal{D}$ is the single data written by the kernel:
\begin{itemize}
  \item $w$ is the single quantity modified by the computation kernel,
  \item $D$ is the computation domain on which $w$ is computed, $D \subseteq entity(w)$,
\end{itemize}
\item $R \in \Delta \times \mathcal{N}$ is the set of tuples $(r,n)$ representing the data read where
\begin{itemize}
  \item $r$ is a quantity read by the kernel to compute $w$,
  \item $n : entity(w) \rightarrow entity(r)^m$ is a neighborhood function that indicates whose values of $r$ are read to compute $w$.
\end{itemize}
\item $comp$ is the numerical computation which returns a value from a set of $n$ input values, $comp: V^n \rightarrow V$, where $V$ is typically $\mathbb{N}$, $\mathbb{Z}$, or $\mathbb{R}$. $comp$ represents the actual numerical expression which is computed by a kernel.
\end{itemize}
\end{mydef}

At each time-step, a set of computations is performed. During a computation kernel, it can be considered that a set of old states ($t-1$) of quantities are read ($R$), and that a new state ($t$) of a single quantity is written ($w$).
%TODO[jb] introduire la contrainte plus tard

Such a definition of a computation kernel covers a large panel of different computations. For example, the four usual types of computations (stencil, local, boundary and reduction) performed into a simulation can be defined as follow :
\begin{itemize}
\item A kernel computation $k(S,R,(w,D),comp)$ is a \emph{stencil kernel} if $\exists (r,n) \in R$ such that $n \neq identity$ .
\item A \emph{boundary kernel} is a kernel $k(S,R,(w,D),comp)$ where $D$ is a specific computation domain at the border of entities, and which does not intersect with any other computation domain.
\item A kernel computation $k(S,R,(w,D),comp)$ is a \emph{local kernel} if $\forall (r,n) \in R$, $n = identity$.
\item A kernel computation $k(S,R,(w,D),comp)$ is a \emph{reduction kernel} if $w$ is a scalar.%, $D=\emptyset$, $R=\{(r,n)\}$ with $n=entity(r)$, and $comp$ is a reduction operation (usually $+$, $-$, $*$, $max$, $min$ etc.).
\end{itemize}
Since we only consider explicit numerical schemes in this paper, a kernel can not write the same quantity it reads, except locally, \textit{i.e.} $(w,n) \in R\Rightarrow n=identity$

%In this paper only explicit numerical schemes are considered, thus if we denote by $identity$ the identity function, we have $w \not\in R$ except if $w=(r,identity)$.

%In our formalism, it is forbidden for a stencil kernel to find $(r,n) \in R$ for which $r=w$ and $n \neq identity$. Actually, if such a property was permitted, implicit numerical schemes would be needed into the simulation, which involves linear solvers. Such a scheme is not a stencil and is over the scope of this paper. The problem does not happened for local kernels as $\forall (r,n) \in R$, $n = identity$. Computations for which this property is authorized are boundary kernels only. This type of kernel is a particular case as it commputes boundary conditions, which represents what should happened outside the space domain but still impact stencil kernels at the next time-step.

@@

% This section has presented the general formalism of what is a computation kernel and a multi-stencil program.
% The language presented in this paper, the Multi-Stencil Language, offers a simple way to describe a numerical simulation while being able to automatically extract parallelism.
% Its aim is to be mesh-agnostic, as previously described, but also to make a clear separation of concerns between the simulation description and its implementation.
% As a result, and as already described, details about the topology of the mesh, which is an implementation problem, is not used into the language.
% Moreover, in a computation kernel $k(S,R,(w,D),comp)$, $comp$ is not used by the language.
